Effects of bariatric surgery on functional connectivity of the reward and default mode network: a pre-registered analysis

### 2021-03-22 manuscript/draft_rsfmri.Rmd
final bib + corrections

<!-- MODEL SETUP --> 
### Confirmatory analyses
**Whole-brain analysis**
We tested the pre-registered hypothesis of a time-by-group interaction:  `FC = group + time + group * time`.\
As pre-registered, we performed the analysis for baseline and followup time points only, for AROMA+CC and AROMA+CC+GSR denoising pipelines and adjusting for no confounders (model CA1), age, sex and average logmFD of both timepoints (model CA2) and age, sex, average logmFD and baseline BMI (model CA3). 
Because the information about the BMI at baseline of one participant in the intervention group was missing, we employed multivariate imputation using the package `mice` based on subject identification, condition, baseline age, and available BMI values. This package allows taking into account the multi-level structure of the data and averaging across multiple possible values, in this case 50 different values, each derived from 10 iterations.\
**Aggregated FC analysis**
We analyzed the aggregate FC using linear mixed models in the R package lme4 [@Bates_2007]. We deviated from the pre-registration by only investigating data from the AROMA+CC+GSR denoising pipeline. First, we investigated the time-by-group interaction for baseline and followup time points only. We either performed no correction (model CA1), adjusted for baseline age, sex, average of logmFD of both time points (model CA2) or additionally adjusted for baseline BMI (model CA3). We performed model comparison between R1 and R0 models, where `R1 = lmer(aggFC ~ timepoint*group + age + sex + (1|subj))`, and `R0 = lmer(aggFC ~ timepoint + group +  age + sex + (1|subj))`. As specified in the pre-registration, we repeated the above-mentioned interaction analysis for all three time points.

@@

```{r figaggFC, fig.cap="Mean network connectivity per group over time separately for the bariatric surgery group (BARS) and the waiting-list control group (NBARS); individual trajectories are plotted in transparent, mean trajectories including standard deviations in opaque colors.", fig.subcap = c("for the DMN", "for the RN"), fig.hold=TRUE, out.width = "50%", fig.align="center", echo = FALSE}
fig_DMNconn<-readRDS(file = "../report/fig/figDMNconn.rds")
fig_Rewconn<-readRDS(file = "../report/fig/figRewconn.rds")
   fig_DMNconn
fig_Rewconn
```
There was no significant group-by-time interaction for aggregated DMN and RN FC (see Figure \@ref(fig:figaggFC) and SI for a detailed summary of the models), regardless of the adjustments (CA1 without adjustment, CA2 adjusting for age,sex and average of mFD or CA3 age,sex, average of mFD and baseline BMI), and whether the analyses were performed in two and three timepoints.

@@

### Average and change in logmFD 
<!-- #### Reward -->As shown in Table \@ref(tab:tableFDmodel) and Figure \@ref(fig:FDagesexNACCccavgFDact), higher average logmFD was associated with stronger FC between the NAcc and a cluster located in proximity to the central sulcus and motor areas. This Cluster only differed in size between denoising pipelines.\
We did not find any clusters with a significant association of logmFD and DMN FC.
```{r "FDagesexNACCccavgFDact", fig.show = "hold", out.width = "50%", fig.cap = "Positive association of average logmFD and FC of NAcc with a cluster in motor cortex, adjusted for age, sex, average BMI, change BMI and change logmFD. Legend denotes empirical Z-values.", fig.subcap = c("in AROMA+CC", "in AROMA+CC+GSR data")}
knitr::include_graphics("../report/fig/FDagesex_NACCcc_avgFD_act.pdf")
knitr::include_graphics("../report/fig/FDagesex_NACCgsr_avgFD_act.pdf")
```
### Functional decoding
Functional decoding of the two activation patterns from the pre-registered, exploratory analysis EA3 showed different top association terms. For the [https://neurosynth.org/decode/?neurovault=441783](https://neurosynth.org/decode/?neurovault=441783) frontal, anterior insula and inferior frontal were the top terms, while for the [https://neurosynth.org/decode/?neurovault=441784](https://neurosynth.org/decode/?neurovault=441784) primary motor, motor and premotor cortex were the meta-analytic activation maps most similar to this contract. The FC contrasts were thus somewhat distinct, though the decoding method [does not allow](https://www.talyarkoni.org/blog/tag/neurosynth/) to conclude specificity.

@@

Taken together, this prospective well-controlled study did not confirm previous findings claiming strong effects of bariatric surgery on FC of the RN and DMN in obese patients. In our cohort, differential changes in head motion adjustment strongly altered rsfMRI neuroimaging results. We thus recommend to rigorously control head motion at acquisition through online monitoring or prospective motion correction, to acquire longer rsfMRI scans or to investigate brain organization with less motion-prone techniques such as task-based fMRI. Furthermore, to increase validity and power, future studies should include obese control groups, and increase efforts to share and pool valuable patient data into meta-analysis. Finally, pre-registration of concrete and testable hypotheses would help to increase replicability of the field and truely enhance our understanding of the neural underpinnings of altered gut-brain communication after bariatric surgery.

# Code availability
All code is available on github. Preprocessing scripts are here [https://github.com/fBeyer89/ADI_preproc](https://github.com/fBeyer89/ADI_preproc), and analyses scripts are here: [https://github.com/hsx1/adi2_rsfMRI](https://github.com/hsx1/adi2_rsfMRI).

>>>>

<!-- MODEL SETUP --> 
### Confirmatory analyses
**Whole-brain analysis**
We tested the pre-registered hypothesis of a time-by-group interaction:  `FC = group + time + group * time`.\
As pre-registered, we performed the analysis for baseline and followup time points only, for AROMA+CC and AROMA+CC+GSR denoising pipelines and adjusting for no confounders (model CA1), age, sex and average logmFD of both time points (model CA2) and age, sex, average logmFD and baseline BMI (model CA3). 
Because the information about the BMI at baseline of one participant in the intervention group was missing, we employed multivariate imputation using the package `mice` based on subject identification, condition, baseline age, and available BMI values. This package allows taking into account the multi-level structure of the data and averaging across multiple possible values, in this case 50 different values, each derived from 10 iterations.\
**Aggregated FC analysis**
We analyzed the aggregate FC using linear mixed models in the R package lme4 [@Bates_2007]. We deviated from the pre-registration by only investigating data from the AROMA+CC+GSR denoising pipeline. First, we investigated the time-by-group interaction for baseline and followup time points only. We either performed no correction (model CA1), adjusted for baseline age, sex, average of logmFD of both time points (model CA2) or additionally adjusted for baseline BMI (model CA3). We performed model comparison between R1 and R0 models, where `R1 = lmer(aggFC ~ timepoint*group + age + sex + (1|subj))`, and `R0 = lmer(aggFC ~ timepoint + group +  age + sex + (1|subj))`. As specified in the pre-registration, we repeated the above-mentioned interaction analysis for all three time points.

@@

```{r figaggFC, fig.cap="Mean network connectivity per group over time separately for the bariatric surgery group (BARS) and the waiting-list control group (NBARS); individual trajectories are plotted in transparent, mean trajectories including standard deviations in opaque colors.", fig.subcap = c("for the DMN", "for the RN"), fig.hold=TRUE, out.width = "50%", fig.align="center", echo = FALSE}
fig_DMNconn<-readRDS(file = "../report/fig/figDMNconn.rds")
fig_Rewconn<-readRDS(file = "../report/fig/figRewconn.rds")
   fig_DMNconn
fig_Rewconn
```
There was no significant group-by-time interaction for aggregated DMN and RN FC (see Figure \@ref(fig:figaggFC) and SI for a detailed summary of the models), regardless of the adjustments (CA1 without adjustment, CA2 adjusting for age,sex and average of mFD or CA3 age,sex, average of mFD and baseline BMI), and whether the analyses were performed in two and three time points.

@@

### Average and change in logmFD 
<!-- #### Reward -->As shown in Table \@ref(tab:tableFDmodel) and Figure \@ref(fig:FDagesexNACCccavgFDact), higher average logmFD was associated with stronger FC between the NAcc and a cluster located in proximity to the central sulcus and motor areas. This Cluster only differed in size between denoising pipelines.\
We did not find any clusters with a significant association of logmFD and DMN FC.
```{r "FDagesexNACCccavgFDact", fig.show = "hold", out.width = "50%", fig.cap = "Positive association of average logmFD and FC of NAcc with a cluster in motor cortex, adjusted for age, sex, average BMI, change BMI and change logmFD. Legend denotes empirical Z-values.", fig.subcap = c("in AROMA+CC", "in AROMA+CC+GSR data")}
knitr::include_graphics("../report/fig/FDagesex_NACCcc_avgFD_act.pdf")
knitr::include_graphics("../report/fig/FDagesex_NACCgsr_avgFD_act.pdf")
```

### Functional decoding
Functional decoding of the two activation patterns from the pre-registered, exploratory analysis EA3 showed different top association terms. For the [https://neurosynth.org/decode/?neurovault=441783](https://neurosynth.org/decode/?neurovault=441783) frontal, anterior insula and inferior frontal were the top terms, while for the [https://neurosynth.org/decode/?neurovault=441784](https://neurosynth.org/decode/?neurovault=441784) primary motor, motor and premotor cortex were the meta-analytic activation maps most similar to this contract. The FC contrasts were thus somewhat distinct, though the decoding method [does not allow](https://www.talyarkoni.org/blog/tag/neurosynth/) to conclude specificity.

@@

Taken together, this prospective well-controlled study did not confirm previous findings claiming strong effects of bariatric surgery on FC of the RN and DMN in obese patients. In our cohort, differential changes in head motion adjustment strongly altered rsfMRI neuroimaging results. We thus recommend to rigorously control head motion at acquisition through online monitoring or prospective motion correction, to acquire longer rsfMRI scans or to investigate brain organization with less motion-prone techniques such as task-based fMRI. Furthermore, to increase validity and power, future studies should include obese control groups, and increase efforts to share and pool valuable patient data into meta-analysis. Finally, pre-registration of concrete and testable hypotheses would help to increase replicability of the field and truly enhance our understanding of the neural underpinnings of altered gut-brain communication after bariatric surgery.

# Code availability
All code is available on github. Preprocessing scripts can be found under [https://github.com/fBeyer89/ADI_preproc](https://github.com/fBeyer89/ADI_preproc), and analyses scripts can be found under [https://github.com/hsx1/adi2_rsfMRI](https://github.com/hsx1/adi2_rsfMRI).
